---
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```
# Lab journal

- Wouter Zeevat
- Thema 7

## Chapter 3

### 3.1 Loading data into R
The data consists of 9 groups, a Deletion, Duplication and control group. And for each of those IPSC, 1m and 3m.
```{R}
library(GEOquery)
```
```{r}
counts <- read.table('GSE142174_16p11.2_CNVs_autism_organoids_study_counts.csv', sep=',', header=TRUE, row.names = 1)
metadata <- getGEO(filename = 'GSE142174_series_matrix.txt')
names(counts) <- metadata@phenoData@data$title
ipsc_control <- 1:12
ipsc_del <- 13:24
ipsc_dup <- 25:36
one_m_control <- 37:48
one_m_del <- 49:60
one_m_dup <- 61:72
three_m_control <- 73:84
three_m_del <- 85:96
three_m_dup <- 97:108
```

### 3.2 Example data
### 3.3 Visualizing using boxplot and density plot
```{R}
library(pander)
```
```{R}
library(scales)
myColors <- hue_pal()(3)
boxplot(log2(counts + 1), outline=FALSE, las=2, col=rep(myColors, each=12), main='Boxplots of all data')

```
This boxplot shows all values, there are not a lot of patterns to see but it looks like the data is not corrupt.
```{R}
library(affy)

myColors <- hue_pal()(3)

## Plot the log2-transformed data with a 0.1 pseudocount
plotDensity(log2(counts + 0.1), col=rep(myColors, each=12),
            lty=c(1:ncol(counts)), xlab='Log2(count)',
            main='Expression Distribution')

## Add a legend and vertical line
abline(v=-1.5, lwd=1, col='red', lty=2)
```
This graph shows the expression distribution. There's a few ones sticking out but its mostly normal.


```{R}
barplot(colSums(counts) / 1e6, col=rep(myColors, each=12), las=2, main='Read counts', ylab='sequence depth')
abline(h = mean(colSums(counts) / 1e6), col = "Black", lty = 5)
```
### 3.4 Sample distance using a heatmap
```{R}
library(pheatmap)
library(DESeq2)

(ddsMat <- DESeqDataSetFromMatrix(countData = counts,
                                  colData = data.frame(samples = names(counts)),
                                  design = ~ 1))

rld.dds <- vst(ddsMat)
# 'Extract' normalized values
rld <- assay(rld.dds)
sampledists <- dist( t( rld ))

# Convert the 'dist' object into a matrix for creating a heatmap
sampleDistMatrix <- as.matrix(sampledists)

# The annotation is an extra layer that will be plotted above the heatmap columns
annotation <- data.frame(Time = factor(rep(1:3, each = 36), 
                                          labels = c("IPSC", "1M", "3M")),
                         Type = factor(rep(rep(1:3, each = 12), 3), 
                                          labels = c("Control", "Del", "Dup")))
                         
# Set the rownames of the annotation dataframe  to the sample names (required)
rownames(annotation) <- names(counts)

pheatmap(sampleDistMatrix, show_colnames = FALSE,
         annotation_col = annotation,
         clustering_distance_rows = sampledists,
         clustering_distance_cols = sampledists,
         main = "Euclidean Sample Distances")
```
Like in the example, the samples are clustered together pretty well. 

### 3.4.4 Multi-Dimensional Scaling
```{R}
library('PoiClaClu')
# Note: uses the raw-count data, PoissonDistance performs normalization
# set by the 'type' parameter (uses DESeq)
dds <- assay(ddsMat)
poisd <- PoissonDistance( t(dds), type = "deseq")
# Extract the matrix with distances
samplePoisDistMatrix <- as.matrix(poisd$dd)
# Calculate the MDS and get the X- and Y-coordinates
mdsPoisData <- data.frame( cmdscale(samplePoisDistMatrix) )

# And set some better readable names for the columns
names(mdsPoisData) <- c('x_coord', 'y_coord')


# Separate the annotation factor (as the variable name is used as label)
groups <- factor(rep(1:9, each=12), 
                 labels = c("IPSC_CONTROL", "IPSC_DEL", "IPSC_DUP", "1M_CONTROL", "1M_DEL", "1M_DUP", "3M_CONTROL", "3M_DEL", "3M_DUP"))
coldata <- names(counts)
```
```{r}
# Create the plot using ggplot
library(ggplot2)
ggplot(mdsPoisData, aes(x_coord, y_coord, color = groups, label = coldata)) + 
  geom_text(size = 4) +
  ggtitle('Multi Dimensional Scaling') +
  labs(x = "Poisson Distance", y = "Poisson Distance") +
  theme_bw()
```
This graph shows the outliers very well. In this case, there aren'y many that differ from the others too much. This is why i decided to not remove them as they can still be crucial to the result.

## Chapter 4
### 4.1 Preprocessing
```{r}
# Perform a naive FPM normalization
# Note: log transformation includes a pseudocount of 1
counts.fpm <- log2( (counts / (colSums(counts) / 1e6)) + 1 )
```
The data will now be normalized by calculating the FPM. 

```{r}
counts.filtered <-  counts.fpm[rowSums(counts.fpm != 0) > 0, ]
counts <-  counts[rowSums(counts.fpm != 0) > 0, ]
```
I decided to remove the rows where all colums are 0.
### 4.2 the fold change value
```{R}
counts.filtered$mean_ipsc_control <- rowMeans(counts.filtered[,ipsc_control])
counts.filtered$mean_ipsc_dup <- rowMeans(counts.filtered[,ipsc_dup])
counts.filtered$fold_change <- counts.filtered$mean_ipsc_dup - counts.filtered$mean_ipsc_control
hist(counts.filtered$fold_change, breaks = 60, main="Log fold changes (IPSC control vs IPSC duplication")
abline(v=-1, col="red")
abline(v=1, col="red")

```
This graph shows that there's some outliers which means that there are probably some DEGs.

### 4.3 Using Bioconductor Packages

### 4.3.1 The design (matrix)
```{r}
library(DESeq2)

(design <- model.matrix(~0 + group))
colnames(design) <- levels(group)
dds <- DESeqDataSetFromMatrix(countData = counts, colData = data.frame(group = group),
                              design = ~ 0 + group)
dds <- DESeq(dds, betaPrior = FALSE)
rn <- resultsNames(dds)
```

### 4.3.2 DESeq
```{r}
result.1m_control_del = results(dds, contrast = c("group", "1M_Control", "1M_Del"), alpha = 0.05)
result.1m_control_dup = results(dds, contrast = c("group", "1M_Control", "1M_Dup"), alpha = 0.05)
result.1m_del_dup = results(dds, contrast = c("group", "1M_Del", "1M_Dup"), alpha = 0.05)
```
```{r}
plotMA(result.1m_control_del)
plotMA(result.1m_control_dup)
plotMA(result.1m_del_dup)
```

### 5.1 Vulcano plot
```{r}
library("EnhancedVolcano")
deseq.volcano <- function(res, datasetName) {
  return(EnhancedVolcano(res, x = 'log2FoldChange', y = 'padj',
                         lab=rownames(res),
                         title = paste(datasetName, "trisomic vs disomic"),
                         subtitle = bquote(italic('FDR <= 0.05 and absolute FC >= 2')),
                         # Change text and icon sizes
                         labSize = 3, pointSize = 1.5, axisLabSize=10, titleLabSize=12,
                         subtitleLabSize=8, captionLabSize=10,
                         # Disable legend
                         legendPosition = "none",
                         # Set cutoffs
                         pCutoff = 0.05, FCcutoff = 2))
}
deseq.volcano(res = result.1m_control_del, datasetName = "IPSC control Del |")
deseq.volcano(res = result.1m_control_dup, datasetName = "IPSC control Dup |")
deseq.volcano(res = result.1m_del_dup, datasetName = "IPSC Del Dup |")
```
